---
title: "STAT 665 HW #3"
author: "Maggie Buffum"
date: "March 7, 2019"
output: pdf_document
---

```{r setup, include=FALSE}
library(pracma)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

## Problem 6.2
Prove the identities (6.1.4):
\[
\begin{aligned}
\bar{\mathbf{x}}_{m+1} &&=&&\frac{1}{m+1}(m\bar{\mathbf{x}}_m + \mathbf{x}_{m+1}),\ \ \ \ \ \ \ \ \ \ \ &\text{ and}&\\
\mathbf{S}_{m+1} &&=&&\mathbf{S}_m + \frac{m}{m+1}(\mathbf{x}_{m+1} - \bar{\mathbf{x}}_m)(\mathbf{x}_{m+1} - \bar{\mathbf{x}}_m)'
\end{aligned}
\]
  
---
  
To show that $\bar x_m = \frac{1}{m+1}(m\bar{\mathbf{x}}_m + \mathbf{x}_{m+1})$, we first recall that $\bar x_m = \frac{1}{m}\sum_{i=1}^mx_i$. Consider $\bar x_{m+1}$:
\[
\begin{aligned}
\bar x_{m+1} &= \frac{1}{m+1}\sum_{i=1}^{m+1}x_i\\
&=\frac{1}{m+1}\left(\sum_{i=1}^{m}x_i + x_{m+1}\right)\\
&=\frac{1}{m+1}\left(m\bar x_m + x_{m+1}\right)\\
\end{aligned}
\]
  
Now let's show that $\mathbf{S}_{m+1} =\mathbf{S}_m + \frac{m}{m+1}(\mathbf{x}_{m+1} - \bar{\mathbf{x}}_m)(\mathbf{x}_{m+1} - \bar{\mathbf{x}}_m)'$:
\[
\begin{aligned}
\mathbf{S}_{m+1} &= \sum_{i=1}^{m+1}(x_i - \bar x_{m+1})(x_i - \bar x_{m+1})'\\
&=\sum_{i=1}^m(x_i - \bar x_{m+1})(x_i - \bar x_{m+1})' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&=\sum_{i=1}^m(x_i - \bar x_m + \bar x_m - \bar x_{m+1})(x_i - \bar x_m + \bar x_m - \bar x_{m+1})' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&=\sum_{i=1}^m\left[(x_i - \bar x_m) + (\bar x_m - \bar x_{m+1})\right]\left[(x_i - \bar x_m) + (\bar x_m - \bar x_{m+1})\right]' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&=\sum_{i=1}^m\left[(x_i - \bar x_m) + (\bar x_m - \bar x_{m+1})\right]\left[(x_i - \bar x_m)' + (\bar x_m - \bar x_{m+1})'\right] + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&=\sum_{i=1}^m(x_i - \bar x_m)(x_i - \bar x_m)' + \sum_{i=1}^m(x_i - \bar x_m)(\bar x_m - \bar x_{m+1})' + \sum_{i=1}^m(\bar x_m - \bar x_{m+1})\left[(x_i - \bar x_m)' + (\bar x_m - \bar x_{m+1})'\right]\\&\ \ \ \ \  + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&= \mathbf{S}_m + \left[\sum_{i=1}^m(x_i - \bar x_m)\right](\bar x_m - \bar x_{m+1})' + \sum_{i=1}^m(\bar x_m - \bar x_{m+1})(x_i - \bar x_{m+1})' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&= \mathbf{S}_m + (m\bar x_m - m\bar x_m)(\bar x_m - \bar x_{m+1})' + (\bar x_m - \bar x_{m+1})\sum_{i=1}^m(x_i - \bar x_{m+1})' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&= \mathbf{S}_m + (\bar x_m - \bar x_{m+1})(m\bar x_m - m\bar x_{m+1})' + (x_{m+1} - \bar x_{m+1})(x_{m+1} - \bar x_{m+1})'\\
&= \mathbf{S}_m + m(\bar x_m - \frac{1}{m+1}(m\bar x_m + x_{m+1}))(\bar x_m - \frac{1}{m+1}(m\bar x_m + x_{m+1}))' \\&\ \ \ \ \ + (x_{m+1} - \frac{1}{m+1}(m\bar x_m + x_{m+1}))(x_{m+1} - \frac{1}{m+1}(m\bar x_m + x_{m+1}))'\\
&= \mathbf{S}_m + \frac{m}{(m+1)^2}(\bar x_m - x_{m+1})(\bar x_m - x_{m+1})' + \frac{m^2}{(m+1)^2}(x_{m+1} - \bar x_m)(x_{m+1} - \bar x_m)'\\
&= \mathbf{S}_m + \frac{m}{(m+1)^2}(-1)^2(x_{m+1}-\bar x_m)(x_{m+1}-\bar x_m)' + \frac{m^2}{(m+1)^2}(x_{m+1} - \bar x_m)(x_{m+1} - \bar x_m)'\\
&= \mathbf{S}_m + \frac{m + m^2}{(m+1)^2}(x_{m+1}-\bar x_m)(x_{m+1}-\bar x_m)'\\
&= \mathbf{S}_m + \frac{m}{m+1}(x_{m+1}-\bar x_m)(x_{m+1}-\bar x_m)'\\
\end{aligned}
\]
  
## Problem 6.4
Suppose $\mathbf{S}_{N_1} \sim W_k(\Sigma,N_1-1)$ and is independent of $\mathbf{S}_{N_2} \sim W_k(\Sigma,N_2 - 1)$. For any $\mathbf{a} = (a_1,a_2,\dots,a_k)'$, find the distribution of $\mathbf{a}'\mathbf{S}_{N_1}\mathbf{a}/\mathbf{a}'\mathbf{S}_{N_2}\mathbf{a}$.
  
---
  
From Example 6.1.1 we have that $\mathbf{a}'\mathbf{S}_{N}\mathbf{a}/\mathbf{a}'\Sigma \mathbf{a} \sim \chi^2_{N-1}$. Note that since the first parameter in each wishart random variable, we have that
\[
\frac{\mathbf{a}'\mathbf{S}_{N_1}\mathbf{a}}{\mathbf{a}'\mathbf{S}_{N_2}\mathbf{a}} = \frac{\mathbf{a}'\mathbf{S}_{N_1}\mathbf{a}/\mathbf{a}'\Sigma \mathbf{a}}{\mathbf{a}'\mathbf{S}_{N_2}\mathbf{a}/\mathbf{a}'\Sigma \mathbf{a}}
\]
  
is now a ratio of two chi-squared random variables. Consider multiplying and dividing each chi-squared random variable by its degrees of freedom. Then we have a scaled F-distribution:
  
\[
\begin{aligned}
&=\frac{(N_1 - 1)}{(N_2 - 1)}\frac{\mathbf{a}'\mathbf{S}_{N_1}\mathbf{a}/(N_1 - 1)}{\mathbf{a}'\mathbf{S}_{N_2}\mathbf{a}/(N_2 - 1)} \sim \frac{(N_1 - 1)}{(N_2 - 1)}F_{(N_1 - 1,N_2-1)}\\
\end{aligned}
\]
  
## Problem 6.6
The following table gives the means and the variance-covariance matrix of four variables $X_0$, $X_1$, $X_2$, and $X_3$:
  
```{r,echo = F}
cov_mat <- matrix(c(
  60.516,0.998,3.511,21.122,
  0.998,15.129,23.860,1.793,
  3.511,23.860,54.756,3.633,
  21.122,1.793,3.633,18.225
),nrow = 4,dimnames = list(c("X_0","X_1","X_2","X_3"),
                           c("X_0","X_1","X_2","X_3")))

means <- matrix(c(
  18.3,14.9,30.5,7.8
),nrow = 4, dimnames = list(c("X_0","X_1","X_2","X_3"),"Means"))

kable(cbind(cov_mat,means),align = "c")
```
  
(a) Compute the multiple correlation coefficient of $X_0$ on $X_1,X_2$ and $X_3$.
  
---
  
Given that in part (b) we want to find the partial correlation between $X_0$ and $X_3$, I'm going to rearrange the variance-covariance matrix as
  
```{r,echo = F}
cov_mat <- matrix(c(
  60.516,21.122,3.511 ,0.998 ,
  21.122,18.225,3.633 ,1.793 ,
  3.511 ,3.633 ,54.756,23.860,
  0.998 ,1.793 ,23.860,15.129
),nrow = 4,dimnames = list(c("X_0","X_3","X_2","X_1"),
                           c("X_0","X_3","X_2","X_1")))

kable(cov_mat,align = "c")
```
  
Let's partition the matrix such that
  
```{r}
S_00 <- as.matrix(cov_mat[1,1])
s_01 <- t(as.matrix(cov_mat[1,2:4]))
s_10 <- as.matrix(cov_mat[2:4,1])
S_1  <- as.matrix(cov_mat[2:4,2:4])

S_00
s_01
s_10
S_1
```
  
Using Result 6.2.5, we can calculate the $\hat\rho_{0(1,\dots,k)}$ as
  
```{r}
rho_hat <- (1/(S_00)^(1/2)) %*% (s_01 %*% solve(S_1) %*% s_10)^(1/2)
rho_hat
```
  
(b) Compute the partial correlation coefficient between $X_0$ and $X_3$, eliminating the effects of $X_1$ and $X_2$.
  
---
  
Consider now that we partition the variance-covariance matrix as
\[
\left[
\begin{array}{cc|cc}
60.516&21.122&3.511 &0.998 \\
21.122&18.225&3.633 &1.793 \\
\hline
3.511 &3.633 &54.756&23.860\\
0.998 &1.793 &23.860&15.129
\end{array}
\right]
\]
  
Using Result 6.2.4 we can calculate $\hat\rho_{jl}|(q+1,\dots,k)$ as
\[
\frac{|\mathbf{S}_{11.2}|_{j,l}}{|\mathbf{S}_{11.2}|_{j,j}^{1/2}|\mathbf{S}_{11.2}|_{l,l}^{1/2}}
\]
  
where
  
```{r}
S_11 <- as.matrix(cov_mat[1:2,1:2])
S_12 <- as.matrix(cov_mat[1:2,3:4])
S_21 <- as.matrix(cov_mat[3:4,1:2])
S_22 <- as.matrix(cov_mat[3:4,3:4])

S_11
S_12
S_21
S_22
```
  
such that
  
```{r}
S_11.2 <- S_11 - S_12 %*% solve(S_22) %*% S_21  
S_11.2
```
  
and
   
```{r}
rho_hat <- S_11.2[1,2] %*% 1/(S_11.2[1,1]^(1/2) %*% S_11.2[2,2]^(1/2))
rho_hat
```
  
## Problem 6.7
Suppose $\mathbf{S}$ is partitioned as in (6.2.9). Evaluate $\hat\rho_{0(1,\dots,k)}^2$ in terms of $|\mathbf{S}|$, $|\mathbf{S}^{(1)}|$ and $S_{00}$.
  
---
  
From (6.2.9), we have that $\mathbf{S}$ is partitioned as
\[
\mathbf{S} = 
\left(
\begin{array}{cc}
S_{00} & \mathbf{s}_{01}\\
\mathbf{s}_{10} & \mathbf{S}^{(1)}
\end{array}
\right)
\]
  
By Result 6.2.4, we can estimate the multiple correlation coefficient $\hat\rho_{0(1,\dots,k)}$ as
\[
\frac{\left(\mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right)^{1/2}}{S_{00}^{1/2}}
\]
  
Let's solve for $\hat\rho_{0(1,\dots,k)}^2$.
\[
\begin{aligned}
\hat\rho_{0(1,\dots,k)}^2 &=\left[ \frac{\left(\mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right)^{1/2}}{S_{00}^{1/2}}\right]^2\\
&= \frac{\mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}}{S_{00}}\\
&= \frac{S_{00} - S_{00} + \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}}{S_{00}}\\
&= \frac{S_{00} - \left[S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right]}{S_{00}}\\
&= \frac{S_{00} - |\mathbf{S}^{(1)}|^{-1}|\mathbf{S}^{(1)}|\left[S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right]}{S_{00}}\\
\end{aligned}
\]
  
Since $S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}$ is a $1\times 1$ matrix, it is its determinant, i.e., $(S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}) = |S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}|$ such that we have
\[
\begin{aligned}
\hat\rho_{0(1,\dots,k)}^2 &= \frac{S_{00} - |\mathbf{S}^{(1)}|^{-1}|\mathbf{S}^{(1)}|\left[S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right]}{S_{00}}\\
&= \frac{S_{00} - |\mathbf{S}^{(1)}|^{-1}|\mathbf{S}^{(1)}|\left|S_{00} - \mathbf{s}_{01}[\mathbf{S}^{(1)}]^{-1}\mathbf{s}_{10}\right|}{S_{00}}\\
&= \frac{S_{00} - |\mathbf{S}^{(1)}|^{-1}|\mathbf{S}|}{S_{00}}\\
&= 1 - \frac{|\mathbf{S}^{(1)}|^{-1}|\mathbf{S}|}{S_{00}}\\
&= 1 - \frac{|\mathbf{S}|}{S_{00}|\mathbf{S}^{(1)}|}\\
\end{aligned}
\]
  
## Problem 6.8
When $\rho_{0(1,\dots,k)}=0$, show that the statistic $\hat\rho^2_{0(1,\dots,k)}$ follow a Beta distribution. What are the parameters of the distribution?
  
---
  
Let $\rho_{0(1,\dots,k)}=0$. Then, from Result 6.2.5 we have that
\[
\frac{(N-k-1)\hat\rho_{0(1,\dots,k)}^2}{k(1 - \hat\rho_{0(1,\dots,k)}^2)} \sim F_{(k,N-k-1)}
\]
  
From Appendix 14, we know that if $V\sim F_{p,q}$ then
\[
X = \frac{(p/q)V}{1 + (p/q)V}\sim Beta(\frac{p}{2},\frac{q}{2})
\]
  
so let's consider $\frac{p}{q}V$, where $V = \hat\rho_{0(1,\dots,k)}^2$, $p = k$, and $q=N-k-1$, such that we have
\[
\frac{p}{q}V = \frac{k}{N-k-1}\left(\frac{(N-k-1)\hat\rho_{0(1,\dots,k)}^2}{k(1-\hat\rho_{0(1,\dots,k)}^2)}\right) = \frac{\hat\rho_{0(1,\dots,k)}^2)}{1-\hat\rho_{0(1,\dots,k)}^2)}
\]
  
Using Appendix 14, we can calculate $X$ as
\[
\begin{aligned}
X &= \frac{\frac{\hat\rho_{0(1,\dots,k)}^2}{1-\hat\rho_{0(1,\dots,k)}^2}}{1 + \frac{\hat\rho_{0(1,\dots,k)}^2}{1-\hat\rho_{0(1,\dots,k)}^2}}\\
&=\frac{\frac{\hat\rho_{0(1,\dots,k)}^2}{1-\hat\rho_{0(1,\dots,k)}^2}}{\frac{1-\hat\rho_{0(1,\dots,k)}^2 +\hat\rho_{0(1,\dots,k)}^2}{1-\hat\rho_{0(1,\dots,k)}^2}}\\
&=\frac{\frac{\hat\rho_{0(1,\dots,k)}^2}{1-\hat\rho_{0(1,\dots,k)}^2}}{\frac{1}{1-\hat\rho_{0(1,\dots,k)}^2}}\\
&=\frac{\hat\rho_{0(1,\dots,k)}^2(1-\hat\rho_{0(1,\dots,k)}^2)}{1-\hat\rho_{0(1,\dots,k)}^2}\\
&=\hat\rho_{0(1,\dots,k)}^2
\end{aligned}
\]
  
giving us that $X = \hat\rho_{0(1,\dots,k)}^2 \sim Beta(\frac{p}{2},\frac{q}{2}) = Beta(\frac{k}{2},\frac{N-k-1}{2})$.








































































